---
title: "Environmental covariate datasets"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Background

This folder contains candidate environmental covariates that could be included in the population models.

The global climate indices included are:

* North Atlantic Oscillation<sup>1</sup>
* Atlantic Multidecadal Oscillation<sup>2</sup>
* Subpolar Gyre Index<sup>3</sup>

1. https://www.ncdc.noaa.gov/teleconnections/nao/
2. https://psl.noaa.gov/data/timeseries/AMO/
3. Generated from the second EOF of ssh produced by the ECMWF ORAS5 global reanalysis hindcast simulation  
https://icdc.cen.uni-hamburg.de/en/daten/reanalysis-ocean/easy-init-ocean/ecmwf-oras5.html  
Zuo, H., Balmaseda, M. A., Tietsche, S., Mogensen, K., and Mayer, M.: The ECMWF operational ensemble reanalysis–analysis system for ocean and sea ice: a description of the system and assessment, Ocean Sci., 15, 779–808, https://doi.org/10.5194/os-15-779-2019, 2019.


### Reading in the data

You can load the files as follows:

```{r example, warning = F, message = F}
require(tidyverse)
require(lubridate)

NAO <- read_csv("NAO.csv")
NAO <- NAO %>% mutate(DATE = make_date(YEAR, MONTH))

AMO <- read_csv("AMO.csv")
AMO <- AMO %>% mutate(DATE = make_date(YEAR, MONTH))

SGI <- read_csv("SGI.csv")

```

## Plotting the data

You can also embed plots, for example:

```{r plots, warning = F, message = F}
require(patchwork)

p1 <- ggplot() + geom_point(aes(x = DATE, y = VALUE), data = NAO) + ylab("NAO")
p2 <- ggplot() + geom_point(aes(x = DATE, y = VALUE), data = AMO) + ylab("AMO")
p3 <- ggplot() + geom_point(aes(x = YEAR, y = VALUE), data = SGI) + ylab("SGI")

p1 / p2 / p3
```

